HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-12 scnn.tex
edits to the method

% MATH SYMBOLS
\newcommand{\1}{\b{1}}              % all-ones vector
\newcommand{\0}{\b{0}}              % all-zero vector
\newcommand{\g}[1]{\b{#1}}
\renewcommand{\L}{\b{L}} % the laplacian matrix
\newcommand{\W}{\b{W}}
\newcommand{\I}{\b{I}}
\newcommand{\D}{\b{D}}
\newcommand{\U}{\b{U}}
\newcommand{\bLambda}{\b{\Lambda}}
\newcommand{\blambda}{\b{\lambda}}
\newcommand{\pkg}[1]{\texttt{#1}}

@@

\begin{figure*}[!ht]
	\centering
	\includegraphics[width=\linewidth]{figures/architecture}
	\caption{Overall network architecture, showing here two convolutional layers acting as feature extractors followed by a fully connected layer with softmax acting as the classifier. A convolutional layer is based on five operations: convolution, non-linearity, batch normalization, down-sampling, and pooling. While most operations are agnostic to the data domain, the convolution and the down-sampling have to be adapted. In this paper, we propose first to model the sphere with a graph and to perform the convolution on the graph. Graphs are versatile data structures which can model any sampling, even irregular or partial. Second, we propose to exploit a hierarchical pixelization of the sphere for the pooling operation. It allows the network to analyze the data at multiple scales while preserving the spatial localization of features.
    \nati{Great plot! Could you make the sphere round? If you use the current code in paper plot notebook, they will be.}
    }
\end{figure*}

@@

To overcome those shortcomings, we introduce an efficient spherical CNN formulation which uses (i) convolutions on graphs and (ii) hierarchical pooling. The main idea is to represent the sphere $S^2$, a 2-dimensional manifold embedded in $\mathbb{R}^3$, as a graph of connected pixels.
Our approach uses the graph CNN formulation introduced by \citet{defferrard2016convolutional}, which is efficient and can adapt to any sampling of the sphere. Using a k-nearest neigboors graph, the convolution operation costs $\mathcal{O}(n)$ operations, where $n$ is the number of pixels. This is the lowest possible complexity for a convolution operation without approximations (e.g. by sketching).
The flexibility of modeling the data domain with a graph allows, for example, to easily model data which only spans a part of the sphere, or data which is not uniformly sampled.
Our hierarchical pooling strategy exploits a hierarchical pixelization of the manifold to analyze the data at multiple scales.
% Such hierarchical pixelizations of the sphere have been well studied.
sdfsdfsd Such hierarchical pixelizations of the sphere have been well studied.
%TK: chagned becasue of blahblah

@@

The first approach, leveraging the continuous spherical convolution, has been
proposed by~\cite{cohen2017convolutional,cohen2018spherical}. The goal of
these contributions it to use a rotational equivariant convolution on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The resulting convolution is performed by a spherical Fourier transform (i.e.
a projection on the spherical harmonics), a  multiplication in the spectral
domain and an inverse spherical Fourier transform. Hence the main
computational cost of a convolution comes from the two Fourier transforms.
Fortunately, the Fourier transform can be accelerated for special samplings
set (theoretically including HealPix, \cite{mohlenkamp1999fast,rokhlin2006fast}).
The main advantage of this
approach is that it provides a mathematically well defined rotational
equivariant network. Nevertheless, even with these Fourier transform
acceleration, the convolutions remain expensive in comparison to a traditional
2-dimensional convolution, limiting the practical use of this approach.
\todo{to have numbers to back up our claim: have the theoretical complexities and measure the time needed for a convolution on graph and the healpix Fourier transform to get the constants
\nati{@mdeff: for the complexity we probably have $\mathcal{O}(n)$ instead of $\mathcal{O}(n^{3/2})$ (let assume that we have $\approx n^{1/2}$ points allong both axis, computation of the SO3 FFT is composed of 1) $n^{1/2}$ FFTs for a total cost of $n \log (n)$ and 2) $n^{1/2}$ matrix multiplications of size $n^{1/2}$ for a total cost of $n^{3/2}$) For the number, it is a bit more complicated… I guess it is for the next paper.}
}
While a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the Healpix one. Hence, a practical comparison is
not possible. We also note that this approach cannot be trivially accelerated
if only one part of the sphere is used. This is impractical in the field of
cosmology as the data often observed only on a subpart of the sphere.
One could still pad the unused part of the sphere with 0. However, this would significantly increase the computational cost of the method.
\nati{@mdeff: do you think we can say the method would be too slow for our dataset even without trying it? I am not so sure...}
\todo{slow: full sky maps are composed of millions of points. partial maps are hard to process}

@@

The gist of our method is to define the convolution on a sphere using a graph.
The graph is here seen as a discrete approximation of the sphere $S^2$, a
2-dimensional manifold embedded in $\mathbb{R}^3$.

As presented by~\cite{cohen2018spherical}, the most mathematical approach to
extend the convolution on a sphere is to use a spherical Fourier transform. The
convolution is then simply defined as the product in the spectral domain. This
approach requires one Fourier and one inverse Fourier transform per convolution
which remain, even with accelerated algorithms, expensive. For 2-dimensional
images, an efficient convolution can be achieved when the convolution kernel is
localized (for example a 5x5 pixel patch) by doing the computation directly in
the signal domain. Unfortunately, this approach cannot be directly extended to
the spherical case. Hence, the main idea of this contribution is to leverage
graph signal processing~\cite{shuman2013emerging} to define a spherical
convolution that can be computed (and back-propagated) directly in the signal domain.

@@

\begin{figure}[!ht]
\centering
% \includegraphics[width=0.45\textwidth]{figures/healpix_graph_4.pdf}
\vspace{-0.5cm}
\includegraphics[width=0.35\textwidth]{figures/half_graph_4.pdf}
\vspace{-0.5cm}
\caption{
Top view of a Healpix graph for half the sphere using a HealPix nside parameter of $4$.
Nodes with 7 neighbors are highlighted. \label{fig:healpix_graph_4}
}

\end{figure}

\subsection{Graph Fourier basis and spherical harmonics}
\todo{Add a few extra references}

Because of the irregular sampling, there is no obvious way to define the convolution
directly from the pixel values. Hence we will first make a detour and define a Fourier transform on the
graph.

@@

\subsection{Convolutions on graphs}
\assign{Nathanaël, Michaël} \todo{Add a few extra references}

Graph convolution is defined by generalizing the concept that convolution is a
multiplication in the spectral domain. Given a convolution kernel
$h:\mathbb{R}_+\rightarrow\mathbb{R}$, the convolution of a signal $\b{f}$
is defined as
\begin{equation} \label{eqn:graph_convolution}
h(L)\b{f} = \U h(\bLambda) \U^* \b{f},
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii}=h(\bLambda_{ii})$.
Equation~\eqnref{graph_convolution} contains three parts: a) $\U^* \b{f}$, the
Fourier transform of $\b{f}$; b) $h(\bLambda)$ the multiplication with the Fourier
kernel $h(\blambda)$; and c) $\U$ the inverse Fourier transform.

The major difference with the classical convolution is that the convolution kernel
$h$ cannot be thought as a $n \times n$ pattern, but is a continuous function
applied to the graph eigenvalues $\bLambda$. For visualization purposes, one can
look at the effect of the convolution on a Kronecker delta, i.e., one column of the matrix
$h(\L)$.
Due to the non-regularity of the graph (i.e. the fact
that there is no perfect sampling on the sphere), the convolution effect will slightly
differ from one node to another. In the specific case of the full sphere, these
differences are negligible in most of the cases. However, when considering only parts of
the sphere, one can observe important border effects (See \figref{border_effects} in Appendix~\ref{sec:border_effects}).
\todo{can we show those?\nati{Sure I just added a figure in the Appendix.}}

@@

\subsection{Efficient convolutions}
\assign{Michaël}
With the Fourier basis. Because we don't have an FFT, that costs $O(n^3)$ for the eigendecomposition, plus $O(n^2)$ for the transform (to be done for each forward and backward pass).
\begin{itemize}
	\item We need efficient convolution, hence the Chebysheff trick
	\item Explain the different between a polynomial filter convolution and a traditional patch restricted convolution
	\item Define spherical CNN using graph CNN
\end{itemize}

@@

Hierarchical pixelization schemes present a natural coarsening. Indeed, each subdivision divides a cell in an equal number of sub-cells. Coarsening is the reverse operation: merging the sub-cells to summarize the data supported on them. Merging cells lead to a coarser graph, hence the name coarsening. Pooling refers to the operation that summarizes the data supported on the merged sub-cells in one child cell. That operation is often taken as the maximum value, but it might be any permutation invariant operation such as the sum or the average. \figref{pooling} illustrates the process.

>>>>

% MATH SYMBOLS
\newcommand{\1}{\b{1}}              % all-ones vector
\newcommand{\0}{\b{0}}              % all-zero vector
\newcommand{\g}[1]{\b{#1}}
\renewcommand{\L}{\b{L}} % the laplacian matrix
\newcommand{\W}{\b{W}}
\newcommand{\I}{\b{I}}
\newcommand{\D}{\b{D}}
\newcommand{\U}{\b{U}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\bLambda}{\b{\Lambda}}
\newcommand{\blambda}{\b{\lambda}}
\newcommand{\pkg}[1]{\texttt{#1}}

@@

\begin{figure*}[!ht]
	\centering
	\includegraphics[width=\linewidth]{figures/architecture}
	\caption{Overall network architecture, showing here two convolutional layers acting as feature extractors followed by a fully connected layer with softmax acting as the classifier. A convolutional layer is based on five operations: convolution, non-linearity, batch normalization, down-sampling, and pooling. While most operations are agnostic to the data domain, the convolution and the down-sampling have to be adapted. In this paper, we propose first to model the sphere with a graph and to perform the convolution on the graph. Graphs are versatile data structures which can model any sampling, even irregular or partial. Second, we propose to exploit a hierarchical pixelization of the sphere for the pooling operation. It allows the network to analyze the data at multiple scales while preserving the spatial localization of features.
    \nati{Great plot! Could you make the sphere round? If you use the current code in paper plot notebook, they will be.}
    }
	\label{fig:architecture}
\end{figure*}

@@

To overcome those shortcomings, we introduce an efficient spherical CNN formulation which uses (i) convolutions on graphs and (ii) hierarchical pooling. The main idea is to represent the sphere $S^2$, a 2-dimensional manifold embedded in $\mathbb{R}^3$, as a graph of connected pixels.
Our approach uses the graph CNN formulation introduced by \citet{defferrard2016convolutional}, which is efficient and can adapt to any sampling of the sphere. Using a k-nearest neigboors graph, the convolution operation costs $\mathcal{O}(n)$ operations, where $n$ is the number of pixels. This is the lowest possible complexity for a convolution operation without approximations (e.g. by sketching).
The flexibility of modeling the data domain with a graph allows, for example, to easily model data which only spans a part of the sphere, or data which is not uniformly sampled.
Our hierarchical pooling strategy exploits a hierarchical pixelization of the manifold to analyze the data at multiple scales.
Such hierarchical pixelizations of the sphere have been well studied.

@@

% As presented by~\cite{cohen2018spherical}, the most mathematical approach to extend the convolution on a sphere is to use a spherical Fourier transform. The convolution is then simply defined as the product in the spectral domain. This approach requires one Fourier and one inverse Fourier transform per convolution which remain, even with accelerated algorithms, expensive.

The first approach, leveraging the continuous spherical convolution, has been
proposed by~\cite{cohen2018spherical}. The goal of
these contributions it to use a rotational equivariant convolution on the
sphere, i.e. a rotation of the input implies the same rotation of the output.
The resulting convolution is performed by a spherical Fourier transform (i.e.
a projection on the spherical harmonics), a  multiplication in the spectral
domain and an inverse spherical Fourier transform. Hence the main
computational cost of a convolution comes from the two Fourier transforms.
Fortunately, the Fourier transform can be accelerated for special samplings
set (theoretically including HealPix, \cite{mohlenkamp1999fast,rokhlin2006fast}).
The main advantage of this
approach is that it provides a mathematically well defined rotational
equivariant network. Nevertheless, even with these Fourier transform
acceleration, the convolutions remain expensive in comparison to a traditional
2-dimensional convolution, limiting the practical use of this approach.
\todo{to have numbers to back up our claim: have the theoretical complexities and measure the time needed for a convolution on graph and the healpix Fourier transform to get the constants
\nati{@mdeff: for the complexity we probably have $\mathcal{O}(n)$ instead of $\mathcal{O}(n^{3/2})$ (let assume that we have $\approx n^{1/2}$ points allong both axis, computation of the SO3 FFT is composed of 1) $n^{1/2}$ FFTs for a total cost of $n \log (n)$ and 2) $n^{1/2}$ matrix multiplications of size $n^{1/2}$ for a total cost of $n^{3/2}$) For the number, it is a bit more complicated… I guess it is for the next paper.}
}
While a comparison is theoretically possible, the experiments
of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and
do not provide any code for the Healpix one. Hence, a practical comparison is
not possible. We also note that this approach cannot be trivially accelerated
if only one part of the sphere is used. This is impractical in the field of
cosmology as the data often observed only on a subpart of the sphere.
One could still pad the unused part of the sphere with 0. However, this would significantly increase the computational cost of the method.
\nati{@mdeff: do you think we can say the method would be too slow for our dataset even without trying it? I am not so sure...}
\todo{slow: full sky maps are composed of millions of points. partial maps are hard to process}

@@

\todo{Should we give a more general / high-level description of a CNN?}

A CNN is composed of the following main building blocks~\citep{lecun1998cnn}, illustrated in~\figref{architecture}: (i) a convolution, (ii) a non-linearity, (iii) a down-sampling operation, (iv) a pooling operation, and (v), optionally, some normalization.\footnote{A batch normalization has been shown to help the training.~\citep{ioffe2015batchnorm} We have verified this experimentally in our setting as well.} As operations (ii) and (v) are point-wise, they do not depend on the data domain. The pooling operation is simply a permutation invariant aggregation function which does not need to be adapted either. The convolution and down-sampling operations however need to be generalized from Euclidean domains to the sphere.

On regular Euclidean domains, such as 1-dimensional time series or 2-dimensional images, a convolution can be efficiently implemented by sliding a localized convolution kernel (for example a patch 5x5 pixels) in the signal domain. Likewise, down-sampling can be achieved by taking one pixel out of $n$.
Because of the irregular sampling, there is no straightforward way to define a convolution on the sphere directly in the pixel domain.

We will hence make a detour by the spectrum to perform the convolution.\footnote{Note that we only use the spectrum to define the convolution. The implementation however does not need a direct access to the spectrum, which is much more computationally efficient. See \secref{sec:efficient_convolution}.}
The gist of our method is to define the convolution on a sphere using a graph, and the down-sampling using a hierarchical pixelization of the sphere.
This contribution leverages the emerging field of graph signal processing~\citep{shuman2013emerging} which defines fundamental signal processing tools such as the Fourier transform for signals defined on graphs.

@@

\begin{figure}[!ht]
\centering
% \includegraphics[width=0.45\textwidth]{figures/healpix_graph_4.pdf}
\vspace{-0.5cm}
\includegraphics[width=0.35\textwidth]{figures/half_graph_4.pdf}
\vspace{-0.5cm}
\caption{
Top view of a Healpix graph for half the sphere using a HealPix nside parameter of $4$.
Nodes with 7 neighbors are highlighted. \label{fig:healpix_graph_4}
}
\end{figure}

\subsection{Graph Fourier basis and spherical harmonics}

@@

\subsection{Convolutions on graphs}

A convolution can be defined on graphs by the convolution theorem which states that a convolution in the vertex (or time, or spatial) domain is a
multiplication in the spectral domain. Given a convolution kernel
$h: \R_+ \rightarrow \R$, the convolution of a signal $\b{f}$
is defined as
\begin{equation} \label{eqn:graph_convolution}
h(L)\b{f} = \U h(\bLambda) \U^* \b{f},
\end{equation}
where $h(\bLambda)$ is a diagonal matrix where $h(\bLambda)_{ii}=h(\bLambda_{ii})$.
Equation~\eqnref{graph_convolution} contains three parts: a) $\U^* \b{f}$, the
Fourier transform of $\b{f}$; b) $h(\bLambda)$ the multiplication with the Fourier
kernel $h(\blambda)$; and c) $\U$ the inverse Fourier transform.

The major difference with the classical convolution is that the convolution kernel
$h$ cannot be thought as a $n \times n$ pattern, but is a continuous function
applied to the graph eigenvalues $\bLambda$. For visualization purposes, one can
look at the effect of the convolution on a Kronecker delta, i.e., one column of the matrix
$h(\L)$.
Due to the non-regularity of the graph (i.e. the fact
that there is no perfect sampling on the sphere), the convolution effect will slightly
differ from one node to another. In the specific case of the full sphere, these
differences are negligible in most of the cases. However, when considering only parts of
the sphere, one can observe important border effects (see \figref{border_effects} and~\ref{sec:border_effects}).

@@

\subsection{Efficient convolutions}
\label{sec:efficient_convolution}
\assign{Michaël}

Transition from spectral to spatial domain.

With the Fourier basis. Because we don't have an FFT, that costs $O(n^3)$ for the eigendecomposition, plus $O(n^2)$ for the transform (to be done for each forward and backward pass).
\begin{itemize}
	\item We need efficient convolution, hence the Chebysheff trick
	\item Explain the different between a polynomial filter convolution and a traditional patch restricted convolution
	\item Define spherical CNN using graph CNN
\end{itemize}

@@

Hierarchical pixelization schemes present a natural coarsening. Indeed, each subdivision divides a cell in an equal number of child sub-cells. Coarsening is the reverse operation: merging the sub-cells toward the goal of summarizing the data supported on them. Merging cells lead to a coarser graph, hence the name coarsening.

Pooling refers to the operation that summarizes the data supported on the merged sub-cells in one parent cell. That operation is often taken to be the maximum value, but it might be any permutation invariant operation such as the sum or the average. \figref{pooling} illustrates the process.
